\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% ICDE 2027 Industry & Applications track (single-blind; 12 pages + unlimited refs;
% no appendices; AI-generated-content acknowledgement required).

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage[hidelinks]{hyperref}
\usepackage{textcomp}

\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\newcommand{\arcadedb}{ArcadeDB}

\begin{document}

\title{ArcadeDB: One ACID Engine for Documents, Graphs, Vectors, and Time Series\\
{\large\todo{working title; alternatives in .notes strategy doc}}}

\author{
\IEEEauthorblockN{Taewoon Kim}
\IEEEauthorblockA{HumemAI\\
taewoon@humem.ai}
\and
\IEEEauthorblockN{Roberto Franchini}
\IEEEauthorblockA{Arcade Data\\
r.franchini@arcadedata.com}
\and
\IEEEauthorblockN{Luca Garulli}
\IEEEauthorblockA{Arcade Data\\
l.garulli@arcadedata.com}
}

\maketitle

\begin{abstract}
\todo{Written last. Spine: multi-model unification as a real, adopted system,
one page/WAL/MVCC-transaction pipeline under documents, graphs, key-value,
time series, and vectors; every index type (LSM, full-text, geo, hash,
dense/sparse vector) commits inside the same transaction; Raft replication ships
model-agnostic WAL page diffs; one jar runs embedded in-process or as a server
speaking six wire protocols. Deep dive: what adding the vector model to a unified
substrate takes, including the deliberate transactional-source-of-truth /
eventually-consistent-ANN-graph tradeoff. Evaluation vs specialist engines +
lessons from a decade of multi-model engineering (OrientDB lineage).}
\end{abstract}

\begin{IEEEkeywords}
multi-model databases, ACID transactions, vector search, graph databases,
embedded databases
\end{IEEEkeywords}

\section{Introduction}\label{sec:intro}
\todo{Vector-era motivation: apps hold documents + relationships + embeddings of
the same entities; composed specialist stacks pay consistency/ops glue. Thesis:
unification at the storage layer, not the API gateway. Claims 1--4 from the
audit. Explicitly claim integration, not new algorithms.}

\section{Background and Design Goals}\label{sec:background}
\todo{PROSE PASS NEEDED. Raw co-author input below (Garulli, 2026-07-08).}

ArcadeDB descends from OrientDB, and its design goals are best understood as
deliberate corrections of three OrientDB limitations its authors lived with:
\begin{itemize}
  \item \textbf{Durability.} OrientDB's write-ahead log was asynchronous,
  single-threaded, and (in the authors' assessment) unreliable. ArcadeDB's
  journal is synchronous, parallel, and crash-proof. It follows the point-of-no-return
  discipline detailed in \S\ref{sec:architecture}.
  \item \textbf{Edge storage.} OrientDB managed edges with a tree-bag
  structure; ArcadeDB replaces it with a LIFO linked list that scales
  substantially better at high edge counts.
  \item \textbf{High availability.} The OrientDB multi-master model ``simply
  doesn't work'' (authors' words); ArcadeDB abandons it for a Raft
  leader/follower architecture (\S\ref{sec:ha}).
\end{itemize}
\todo{Design goals framing; what industry usage demanded. Third-party academic
adoption: DatAasee~\cite{himpe2024dataasee} (metadata catalog for research
libraries) and AptBacterialDB~\cite{aptbacterialdb} (ArcadeDB/openCypher graph
layer for an antibacterial-aptamer database), both independent multi-model
graph uses. Christian's "why we chose it" paragraph goes here or in Lessons.}

\section{Architecture: One Substrate, Five Models}\label{sec:architecture}

\arcadedb{} is organized around a single storage abstraction rather than a
federation of per-model engines. Everything that persists, whether it holds
documents, graph adjacency, key-value pairs, time-series points, or vector
index segments, is a \emph{paginated component}: a file of fixed-size pages
managed by one page manager, modified through one transaction context, and
made durable by one write-ahead log. The data models differ in the record
types and index structures layered on top; they do not differ in how bytes
reach disk, how changes become durable, or how replicas stay consistent. This
section walks that pipeline bottom-up and states precisely which guarantees
each model inherits from it.

\subsection{Pages, Transactions, and the Log}\label{sec:pipeline}

All reads and writes go through page-level copy-on-write. A transaction that
modifies any record obtains private copies of the affected pages from the page
manager and accumulates its changes there; concurrent readers continue to see
the unmodified originals. Commit is optimistic: at first phase, the engine
validates that the pages read by the transaction have not been changed by a
concurrent committer (a per-page version check), then serializes every dirty
page, \emph{including index pages}, into a single WAL entry. There is no
separate index log and no index-specific recovery path: a secondary-index
update, a graph edge insertion, and a document field change that happen in the
same transaction land in the same log record and either all survive a crash or
none do. Recovery replays the WAL forward from the last checkpoint; we measure
recovery time and post-crash integrity under \texttt{kill -9} in
\S\ref{sec:eval}.

Durability is configurable per deployment in one dimension only: when the WAL
buffer reaches the operating system (asynchronously on a background flusher by
default, or \texttt{fsync} per commit under the strictest setting). We flag
this knob here because it moves throughput by more than an order of magnitude
and is therefore the first thing our evaluation pins when comparing against
engines with different default contracts (\S\ref{sec:eval}).

\subsection{One Index Family Tree}\label{sec:indexes}

Table~\ref{tab:inventory} inventories the index and storage structures. The
workhorse is an LSM tree~\cite{oneil1996lsm} implemented as a paginated
component: mutable in-memory pages absorb writes, immutable compacted segments
serve reads, and compaction is an ordinary paginated write path. The full-text
index is a thin analyzer layer over the same LSM tree (tokens as keys), and
the geospatial index likewise reduces to LSM entries over space-filling-curve
keys. The practical consequence of this reuse is uniform transactional
behavior: because every index is a paginated component, the commit path of
\S\ref{sec:pipeline} covers all of them with no per-index machinery. The two
vector index families are the newest members of the tree and the most
demanding test of the claim that the substrate generalizes; they get their own
section (\S\ref{sec:vector}).

\begin{table}[t]
\centering
\caption{Storage and index structures. Every row is a paginated component:
same pages, same WAL entry at commit, same replication path. The ANN graph
(marked $\dagger$) is derived state, rebuilt rather than replicated
(\S\ref{sec:vector}).}
\label{tab:inventory}
\small
\begin{tabular}{lll}
\toprule
Structure & Serves & Implementation \\
\midrule
Buckets & documents, vertices, edges, KV & paged record files \\
LSM tree & secondary indexes (unique/non) & mutable + compacted \\
Full-text & text search & analyzer over LSM \\
Geospatial & spatial predicates & curve keys over LSM \\
Adjacency & graph traversal & RID lists, light edges \\
\texttt{LSM\_SPARSE\_VECTOR} & learned sparse retrieval & postings as LSM \\
Vector segments & dense ANN & records + JVector graph$^\dagger$ \\
Time-series buckets & append/range by time & paged, SIMD codecs \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Graph Storage, Stated Honestly}\label{sec:graph}

Vertices and edges are records in buckets, and adjacency is maintained as
RID-linked lists attached to each vertex, a design inherited from
OrientDB~\cite{ritter2021orientdb} and refined (the tree-bag edge container
was replaced by a LIFO linked list that behaves better at high fan-out,
\S\ref{sec:background}). Edges without properties are stored as
\emph{light edges}: pure RID pairs in the adjacency structure with no edge
record at all, which roughly halves the storage and I/O cost of plain
connections. Traversal is pointer-chasing over the record store through the
same page cache as every other access. We do not claim parity with
purpose-built native graph storage on every workload; we claim, and measure in
\S\ref{sec:eval}, that RID-linked adjacency over a shared substrate delivers
competitive point-traversal latency while inheriting transactions, recovery,
and replication that a bolted-on graph layer would have to re-implement.

\subsection{Replication Is Model-Agnostic}\label{sec:replication-arch}

Because every model's changes reduce to WAL'd page deltas, replication does
not know what a document, an edge, or a vector is. The high-availability
module (\S\ref{sec:ha}) ships transaction page deltas through a Raft
log~\cite{ongaro2014raft}; followers apply them through the same recovery
code path used after a crash. Adding a new data model to \arcadedb{}, as
\S\ref{sec:vector} recounts for vectors, required \emph{no} replication work:
a model whose writes are paginated is replicated by construction. The one
deliberate exception, derived search structures that are rebuilt rather than
replicated, is discussed in \S\ref{sec:vector}.

\section{Case Study: Adding the Vector Model}\label{sec:vector}

Vector search is the sharpest recent test of the unification thesis: it
arrived as a workload after the substrate was fixed, it is dominated by
specialist engines with no transactional obligations, and it stresses exactly
the properties (index build cost, memory, approximate results) that a shared
substrate might be suspected of handling poorly. This section describes what
adding dense and sparse vector search to \arcadedb{} actually took, in the
spirit of integration case studies such as
SingleStore-V~\cite{chen2024singlestorev} and Cosmos DB's DiskANN
integration~\cite{upreti2025cosmosvector}, and states the one place where we
deliberately relaxed the substrate's guarantees.

\subsection{Dense Vectors: Transactional Records, Derived Graph}

Dense embeddings are ordinary record properties: they live in buckets, are
WAL'd, versioned, and replicated like any other field. Approximate search is
served by a graph index built with the JVector
library~\cite{jvector2024repo}, which combines an HNSW-style layer
hierarchy~\cite{malkov2020hnsw} with Vamana-style graphs within each
layer~\cite{subramanya2019diskann}. Here we made the one deliberate exception
to full unification: the ANN graph itself is an asynchronously maintained,
WAL-disabled structure. Vector \emph{records} are the transactional source of
truth; the search graph is derived state, rebuilt from records if lost, and
followers rebuild rather than receive it. The alternative, WAL-logging every
neighbor-list mutation of an incremental graph build, would have coupled
commit latency of unrelated transactions to ANN maintenance. The consequence
is a bounded staleness window between a committed insert and its
visibility in approximate search (newly inserted vectors are meanwhile served
exactly from a delta buffer), which we consider the correct trade for an
operational engine and state openly rather than claim a fully transactional
ANN index. One design decision this enabled: because vectors-as-records are
the ground truth, index parameters can be changed and the graph rebuilt
without touching data, which we used in \S\ref{sec:eval} to match graph
degrees fairly across engines.

\subsection{Sparse Vectors: An Inverted Index That Commits}

Learned sparse retrieval (SPLADE-style~\cite{formal2021splade}) maps text to
high-dimensional sparse weight vectors and is conventionally served by
standalone inverted-index engines with block-max
pruning~\cite{broder2003wand,ding2011blockmax,tonellotto2018efficient}. To our
knowledge \arcadedb{} is the first general-purpose transactional DBMS to ship
this workload natively: \texttt{LSM\_SPARSE\_VECTOR} stores posting lists as
LSM segments (the same paginated component family as every other index), with
RID-delta compression, per-block weight quantization to int8, block maxima
for WAND-style pruning at query time, and size-tiered compaction as the
settle step. Because postings ride the LSM commit path, sparse index updates
are transactional with document writes, a property none of the standalone
engines offer. The quantization choice is a measured trade: int8 posting
weights cost about half a point of recall@10 against exact scoring
(\S\ref{sec:eval}) and can be disabled per index (fp32 segments) where exact
weights matter more than space.

\subsection{What Integration Cost, Measured}

Honesty about the bill: the substrate gave vectors durability, transactions,
and replication for free, but not specialist performance for free. Three
costs surfaced in our own measurements (\S\ref{sec:eval},
\S\ref{sec:lessons}): graph index construction at 10M vectors requires several
times the raw vector size in heap because build-time structures hold vectors
on-heap alongside JVector's construction state; sparse query latency on real
SPLADE term distributions degrades at corpus sizes where specialist engines
do not; and settle steps (compaction) that specialists hide behind background
services are visible operations here. We report these as findings with the
same prominence as the wins, and several became upstream fixes during the
course of this work (\S\ref{sec:lessons}).

\subsection{A Note on Measurement Integrity}

The sparse index's original design documentation reported a 766$\times$
speedup over brute-force scoring at 1M documents, a number that survived
review because the same document also \emph{retracted} an earlier
41{,}613$\times$ claim traced to a broken baseline. We mention this not as
trivia but as method: every headline number in this paper carries its
measurement protocol, and where our own instruments were later found
misleading (a memory metric that counted reclaimable page cache; a timeout
that was our harness's fault) the corrected number replaced the flattering
one. \S\ref{sec:eval} states the protocol once, and the raw results are
published with the paper.

\section{Query and Deployment Surfaces}\label{sec:surfaces}
\todo{One executor + result model behind SQL/Cypher/GraphQL/Mongo (Gremlin via
TinkerPop over the same storage); embedded in-process vs server; six wire
protocols; cite the SciPy paper for Python binding internals.}

\section{High Availability}\label{sec:ha}
\todo{Ratis-based Raft shipping shared-WAL page diffs => model-agnostic
replication; quorum/membership hardening lessons.}

\section{Evaluation}\label{sec:eval}

The evaluation asks one question per data model (is \arcadedb{} competitive
with the specialist a practitioner would otherwise deploy?) and one question
no specialist can answer (what does one engine buy when the workload spans
models?). We report losses with the same prominence as wins.

\textbf{Protocol.} All measurements run on a dedicated
Intel i9-12900HK (6 P-cores/20 threads used as pinned cpusets, 61\,GiB RAM,
NVMe SSD, Linux 7.0), every system in Docker with pinned image digests,
identical CPU and memory caps per cell, and JVM heap fixed
(\texttt{-Xms=-Xmx}) for latency stability. Each cell is $N{=}5$ independent
repetitions; tables report median [min--max]. Quality metrics (recall@10
against exact ground truth) are reported next to every approximate-search
latency. Durability contracts differ across engines by default and move
results by orders of magnitude; we state each engine's contract per table and
additionally measure matched-contract pairs. Raw results, harness, and
per-cell manifests are published.\todo{artifact URL; October freeze re-measure
on one pinned release; comparator version list}

\begin{table}[t]
\centering
\caption{Tabular workloads, medium scale + TPC-H SF1. Median [min--max],
$N{=}5$. Durability contracts differ: \arcadedb{} groups commits
(async WAL flush, default); PostgreSQL and DuckDB fsync per commit.
Matched-contract pairing in the text.}
\label{tab:tabular}
\scriptsize
\setlength{\tabcolsep}{3pt}
\input{tables/t2_tabular}
\end{table}

\begin{table}[t]
\centering
\caption{LDBC SNB graph workloads, SF1/SF10. Median [min--max], $N{=}5$.
Bidirectional edge pointers (engine default) for all \arcadedb{} rows.}
\label{tab:graph}
\scriptsize
\setlength{\tabcolsep}{3pt}
\input{tables/t3_graph}
\end{table}

\begin{table}[t]
\centering
\caption{Sparse retrieval on Big-ANN (real SPLADE/MS MARCO), exact MIPS
ground truth, 1000 dev queries. Median [min--max], $N{=}5$. int8 rows use the
engine's default posting quantization; fp32 disables it.}
\label{tab:sparse}
\scriptsize
\setlength{\tabcolsep}{3pt}
\input{tables/t4_sparse}
\end{table}

\begin{table}[t]
\centering
\caption{Dense ANN on DEEP-10M (degree-matched, exact GT) and TSBS cpu-only
time series. Median [min--max], $N{=}5$.}
\label{tab:denses}
\scriptsize
\setlength{\tabcolsep}{3pt}
\input{tables/t5_dense_ts}
\end{table}

\subsection{Tabular: Documents as Rows}
Table~\ref{tab:tabular} (medium scale; TPC-H SF1 for the analytical
queries). \arcadedb{} embedded sustains 5{,}929 OLTP ops/s with 0.54\,ms
insert p99, ahead of PostgreSQL (525 ops/s) and DuckDB (268 ops/s), but the
contracts differ: \arcadedb{} groups commits through an asynchronous WAL
flusher by default while PostgreSQL and DuckDB fsync per commit; under
\arcadedb{}'s strictest per-commit fsync the gap closes to rough parity with
PostgreSQL. The analytical side is an honest loss: link-based row storage
answers TPC-H Q1 in 10.7\,s where columnar DuckDB needs 15.5\,ms. \arcadedb{}
is an operational engine; workloads dominated by full-column scans belong on
a columnar system, and \S\ref{sec:lessons} discusses where that boundary sits.

\subsection{Graph: LDBC Social Network}
Table~\ref{tab:graph} (LDBC SNB persons/knows, SF1 and SF10). Point
traversals are \arcadedb{}'s strongest surface: 1-hop p50 of 0.49\,ms embedded
(0.56\,ms at SF10) versus 4.9\,ms for Neo4j and 1.1\,ms for LadybugDB, with
p99 under 1.5\,ms at both scales. RID-linked adjacency holds its shape as the
graph grows tenfold. The loss column: SF10 2-hop tails favor Neo4j (10.5\,ms
p99 vs.\ our 18.2\,ms), consistent with a planner that has decades of
traversal-ordering work behind it.

\subsection{Dense Vectors: DEEP-10M}
Table~\ref{tab:denses} top half (10M $\times$ 96d, exact GT, graph degrees
matched across engines). \arcadedb{} embedded reaches 0.950 recall@10 at
5.5\,ms p50: mid-pack among eight systems, behind Qdrant (0.980 at 1.3\,ms)
and Chroma, ahead of exact scanners and of LanceDB's operating point (0.707
recall). Index build is the weak axis (2{,}796\,s vs.\ Qdrant's 516\,s) and
its memory profile is dissected in \S\ref{sec:lessons}. We consider mid-pack
latency at matched recall, plus transactions and replication the specialists
lack, the honest summary of where integration stands today.

\subsection{Sparse Vectors: Big-ANN SPLADE}
Table~\ref{tab:sparse} (real SPLADE/MS MARCO vectors, 100k and 1M tiers,
exact MIPS GT). At 100k \arcadedb{} is competitive (1.45\,ms p50, recall
0.996, twice Qdrant's latency). At 1M it is not: p50 degrades to 165\,ms, a
cliff our specialists do not exhibit, which we traced to head-term posting
lists on real term distributions (\S\ref{sec:lessons}; upstream issue filed
during this work). The synthetic corpus used by the engine's own benchmarks
shows no cliff at 10M, which is itself a finding about synthetic sparse
benchmarks. Elasticsearch offers a mirror-image trade at 1M: fast (3.6\,ms)
but at 0.756 recall from aggressive default pruning, versus \arcadedb{}'s
0.994.\todo{update after upstream fix + 8.84M tier at freeze}

\subsection{Time Series: TSBS}
Table~\ref{tab:denses} bottom half. Specialists win ingest by 14--62$\times$
(DuckDB 1.94M pts/s, QuestDB 431k, \arcadedb{} 31k) and \arcadedb{} loses
global aggregation outright (1.8\,s vs.\ milliseconds; no time partitioning).
It wins the operational lookup: last-point-per-host at 0.52\,ms p50 beats
both specialists through the composite (host, ts) index. The lane's purpose
is scope honesty: the substrate handles the shape, and applications with an
embedded time-series side channel get it for free, but a dedicated TSDB it is
not.

\subsection{Cross-Model: One Transaction vs.\ a Composed Stack}
The workload no specialist runs: a vector hit expands over graph edges and
updates a document, atomically. In \arcadedb{} this is one transaction at
3.36\,ms median end-to-end [3.20--3.49]; SurrealDB (the one embeddable
multi-model rival) needs 7.02\,ms; a composed Qdrant+Neo4j stack needs
22.35\,ms of network and glue. Then the failure injection: killing the writer
mid-operation left the composed stack in a torn state in 5 of 5 trials
(vector store and graph disagree on the result), while \arcadedb{} and
SurrealDB rolled back cleanly in 5 of 5. This is the paper's thesis in one
experiment: the composed stack's inconsistency is not an implementation bug
but the absence of a transaction boundary spanning the models.

\subsection{Durability and Failover}
Under \texttt{kill -9} during sustained ingest (600k+ rows in flight),
\arcadedb{} recovers in under one second with zero torn records across five
trials (watermark-verified), both under the default and the strict WAL
contract. A three-node Raft cluster under write load elects and resumes
writes 0.2--3.3\,s after the leader is killed, with all acknowledged writes
present afterward in 5 of 5 trials. Replication required no vector- or
graph-specific code (\S\ref{sec:replication-arch}).

\subsection{Embedded vs.\ Server}
Across lanes, the same engine in-process versus behind HTTP: OLTP 5,929 vs.\
1,288 ops/s, graph 1-hop 0.49 vs.\ 1.25\,ms, dense p50 5.45 vs.\ 6.82\,ms.
The deployment axis costs a fixed per-operation transport fee that dominates
point workloads and vanishes for scan-bound ones; one engine spans both
deployment points, which neither SQLite-class embedded engines nor
server-only specialists offer.\todo{F8 figure}

\section{Lessons Learned and Tradeoffs}\label{sec:lessons}
\todo{The cost of unification (measured); where specialists win; single-node
ceiling; async-ANN consequences; production war stories (co-authors).}

\section{Related Work}\label{sec:related}
\todo{Multi-model (Cosmos DB API-level vs our storage-level; ArangoDB;
Lu \& Holubov\'a survey; UniBench/M2Bench), vector DBs (Milvus, Manu, Qdrant,
pgvector), sparse retrieval (SPLADE, BlockMax-WAND), embedded engines
(SQLite, DuckDB).}

\section{Conclusion}\label{sec:conclusion}
\todo{}

\section*{Acknowledgment}
We thank Christian Himpe (University of M\"unster) for his contributions to
ArcadeDB, for sharing his experience operating it in
DatAasee~\cite{himpe2024dataasee}, and for feedback on this manuscript.
\todo{AI-generated-content disclosure REQUIRED by ICDE 2027: name tools and
affected content areas (drafting/editing assistance, experiment tooling).
Keep updated as the paper evolves.}

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
